\documentclass[twoside]{scrartcl}

\usepackage[blacklinks]{drangreport}
% \usepackage[caecilia]{drangreport}

\usepackage{stmaryrd}

\usepackage{awlc-algebra}
\newcommand*{\bord}{\multicolumn{1}{c|}{}}

%Nice end of Proof Symbol
\usepackage{marvosym}
\renewcommand{\qedsymbol}{\Rectsteel} 

\usepackage{array}

\usepackage{tikz}
\usepackage{tikz-cd}

\usepackage{tkz-euclide,tikz}
\usetkzobj{all}

\def\n{5}
\def\m{4}


\newcommand*{\Range}{%
  \mathsf{X}\mkern-9mu\mathsf{X}%
}

\newcommand*{\Rangey}{%
  \mathsf{Y}\mkern-9mu\mathsf{Y}%
}




\newcommand*\circled[1]{\tikz[baseline=(char.base)]{
            \node[shape=circle,draw,inner sep=2pt] (char) {#1};}}

\usepackage{wrapfig}
\usepackage{permute}

%Smaller Column Vectors
\makeatletter
\newcommand{\Spvek}[2][r]{%
  \gdef\@VORNE{1}
  \left(\hskip-\arraycolsep%
    \begin{array}{#1}\vekSp@lten{#2}\end{array}%
  \hskip-\arraycolsep\right)}

\def\vekSp@lten#1{\xvekSp@lten#1;vekL@stLine;}
\def\vekL@stLine{vekL@stLine}
\def\xvekSp@lten#1;{\def\temp{#1}%
  \ifx\temp\vekL@stLine
  \else
    \ifnum\@VORNE=1\gdef\@VORNE{0}
    \else\@arraycr\fi%
    #1%
    \expandafter\xvekSp@lten
  \fi}
\makeatother

\newenvironment{amatrix}[1]{%
  \left(\begin{array}{@{}*{#1}{c}|c@{}}
}{%
  \end{array}\right)
}


\title{Algebra I}

\let\Im\relax
\DeclareMathOperator*{\Im}{Im}

\newcommand{\herm}[1]{\overline{#1}^\Trans}

\DeclareMathOperator*{\ch}{ch}
\renewcommand{\chi}{\textstyle\ch}

\begin{document}

\NotesTitle{1st}{Spring 2015}{Algebra I}{Prof.~J.}{Britnell}
{


{\Large\bfseries\sffamily{Syllabus}}

\textit{Introductions to three topics in abstract algebra: The theory of vector spaces and linear
transformations, the theory of groups and the theory of rings.}


\begin{itemize}
\item Vector spaces: Linear maps, rank-nullity theorem, connections with linear equations and matrices.

\item Groups: Axioms, examples. Cyclic groups, dihedral groups, symmetric groups. Lagrange’s  theorem and applications.

\item Rings: Polynomial rings, rings of the form Z[$\sqrt{d}$]. Euclid’s algorithm for certain rings. Uniqueness of factorisation for these rings. Applications to Diophantine Equations.
\end{itemize}

\subsection*{Appropriate books}

{\shortskip
J.~Fraleigh and R.~Beauregard, \emph{Linear Algebra}

S.~Lipschutz and M.~Lipson, \emph{Linear Algebra}

J.~B.~Fraleigh, \emph{A First Course in Abstract Algebra}

R.~Allenby, \emph{Rings, Fields and Groups}

I.~N.~Herstein, \emph{Topics in Algebra}
}}

\TableofContents

\input{M1P2-Chap1.tex}		% Vector spaces
\input{M1P2-Chap2.tex}		% Endomorhisms
\input{M1P2-Chap3.tex}		% Endomorhisms
%\input{linalg-chap-03.tex}

\end{document}