\documentclass[twoside]{scrartcl}

\usepackage[blacklinks]{drangreport}
% \usepackage[caecilia]{drangreport}

\def\n{5}
\def\m{4}


\renewcommand{\theequation}{\arabic{equation}}

%Change font of proof to scshape, not sure if it looks better...
\expandafter\let\expandafter\oldproof\csname\string\proof\endcsname
\let\oldendproof\endproof
\renewenvironment{proof}[1][\proofname]{%
  \oldproof[\scshape #1]%
}{\oldendproof}


\hypersetup{pdfinfo={
Title={Algebra I},
Author={Karim Bacchus},
Keywords ={Algebra I, M1P2, Lecture Notes, Imperial College, Maths},
}}


\title{Algebra I}

\begin{document}

\NotesTitle{1st}{Spring 2015}{Algebra I}{Prof.~J.}{Britnell}
{
Caveat Lector: unofficial notes, \emph{not} endorsed by Imperial College.\\[0.1cm]
Comments and corrections should be sent to \href{mailto:kb514@ic.ac.uk}{\texttt{kb514@ic.ac.uk}}
}
{


\subsection*{Syllabus}

\textit{Introductions to three topics in abstract algebra: The theory of vector spaces and linear
transformations, the theory of groups and the theory of rings.}


\begin{itemize}
\item Vector spaces: Linear maps, rank-nullity theorem, connections with linear equations and matrices.

\item Groups: Axioms, examples. Cyclic groups, dihedral groups, symmetric groups. Lagrange’s  theorem and applications.

\item Rings:  Polynomial rings, rings of the form Z[$\sqrt{d}$]. Euclid’s algorithm for certain rings. Uniqueness of factorisation for these rings. Applications to Diophantine Equations.
\end{itemize}

\subsection*{Appropriate books}

{\shortskip
J.~Fraleigh and R.~Beauregard, \emph{Linear Algebra}

S.~Lipschutz and M.~Lipson, \emph{Linear Algebra}

J.~B.~Fraleigh, \emph{A First Course in Abstract Algebra}

R.~Allenby, \emph{Rings, Fields and Groups}

I.~N.~Herstein, \emph{Topics in Algebra}
}}

\TableofContents

\input{M1P2-Chap1.tex}		% Vector Spaces
\input{M1P2-Chap2.tex}		% Group Theory
\input{M1P2-Chap3.tex}		% Ring Theory

\end{document}