\documentclass[twoside]{scrartcl}

\usepackage[blacklinks]{drangreport}
% \usepackage[caecilia]{drangreport}

\usepackage{pgfplots}

%Thomas' contradiction symbol
\newcommand*{\cont}{%
  \mathsf{X}\mkern-8.5mu\mathsf{X}%
}
\renewcommand{\thempfootnote}{$\dagger$}

%Change font of proof to scshape, not sure if it looks better...
\expandafter\let\expandafter\oldproof\csname\string\proof\endcsname
\let\oldendproof\endproof
\renewenvironment{proof}[1][\proofname]{%
  \oldproof[\scshape #1]%
}{\oldendproof}


\title{Analysis I}


\hypersetup{pdfinfo={
Title={Analysis I},
Author={Karim Bacchus},
Keywords ={Analysis I, M1P1, Lecture Notes, Imperial College, Maths},
}}


\begin{document}


\NotesTitle{1st}{Spring 2015}{Analysis I}{Prof.~R.}{Thomas}
{
Caveat Lector: unofficial notes, \emph{not} endorsed by Imperial College.\\[0.1cm]
Comments and corrections should be sent to \href{mailto:kb514@ic.ac.uk}{\texttt{kb514@ic.ac.uk}}
}
{

\section*{About these notes}

These notes are not affiliated with Richard Thomas (or even \emph{Professor} Richard Thomas!)\\ The \LaTeX{} source is available on \texttt{github} - it would be great if someone would update changes to the courses so they'll still be useful to later years. Notes for other courses are available on \texttt{dropbox}:

\begin{center}
\begin{tabular}{l | l}
\textbf{1st / 2nd Year} & \textbf{3rd / 4th Year}\\
Foundations of Analysis & Galois Theory\\
Analysis I & Algebraic Number Theory \\
Analysis II & Analytical Number Theory \\
Complex Analysis & Elliptic Curves  \\
Geometry \& Linear Algebra & Modular Forms \\
Algebra I & Algebra III \\
Algebra II & Commutative Algebra \\
Methods I & Lie Algebras  \\
Methods II & Measure \& Integration  \\
Multivariable Calculus & Functional Analysis\\
Differential Equations & Algebraic Topology \\
Metric Spaces \& Topology  & Differential Topology \\
Intro to Numerical Analysis & Complexity \\
Mechanics & General Relativity \\
\end{tabular}
\end{center}

\begin{flushright}
- Karim Bacchus 	
\end{flushright}



\pagebreak 
\thispagestyle{empty}


\section*{Syllabus}  

\textit{A rigorous treatment of the concept of a limit, as applied to sequences, series and functions.}
\begin{itemize}
\item  Real and complex sequences. Convergence, divergence and divergence to infinity. Sums and products of convergent sequences. The Sandwich Test. Sub-sequences, monotonic sequences, Bolzano-Weierstrass Theorem. Cauchy sequences and the general principle of convergence.

\item Real and complex series. Convergent and absolutely convergent series. The Comparison Test for non-negative series and for absolutely convergent series. The Alternating Series Test.  Rearranging absolutely convergent series. Radius of convergence of power series. The exponential series.

\item Limits and continuity of real and complex functions. Left and right limits and continuity. Maxima
and minima of real valued continuous functions on a closed interval. Inverse Function Theorem for
strictly monotonic real functions on an interval. 

\item  An introduction to differentiability: definitions, examples, left and right derivative.
\end{itemize}
\subsection*{Appropriate books}

{\shortskip

K.~G.~ Binmore, \textit{Mathematical Analysis, A Straightforward Approach} (Cambridge University Press).

}

}





\TableofContents
\input{Chap1.tex}		% Sequences
\input{Chap2.tex}		% Series
\input{Chap3.tex}		% Continuity
\input{Chap4.tex}		% Differentiation


\end{document}