\documentclass[twoside]{scrartcl}

\usepackage[blacklinks]{drangreport}
% \usepackage[caecilia]{drangreport}

\usepackage{stmaryrd}

\usepackage{awlc-algebra}
 \newcommand{\RR}{\ensuremath{\mathbb{R}}}
\usepackage{pgfplots}
%Nice end of Proof Symbol
\usepackage{marvosym}
\renewcommand{\qedsymbol}{\Rectsteel} 

\newcommand*{\cont}{%
  \mathsf{X}\mkern-8.5mu\mathsf{X}%
}
\renewcommand{\thempfootnote}{$\dagger$}

\title{Analysis I}

\let\Im\relax
\DeclareMathOperator*{\Im}{Im}

\newcommand{\herm}[1]{\overline{#1}^\Trans}

\DeclareMathOperator*{\ch}{ch}
\renewcommand{\chi}{\textstyle\ch}

\begin{document}

\NotesTitle{1st}{Spring 2015}{Analysis I}{Prof.~R.}{Thomas}
{

%\setcounter{equation}{-1}
%\begin{theorem}
%$\forall$ lecture notes,  $\exists$ typos.
%\end{theorem}
%\begin{corollary}
%Don't rely on this. Send comments / corrections to \url{kb514@ic.ac.uk}	
%\end{corollary}\vspace*{10pt}



%*If you're bored with a clicker question filled lecture, \texttt{piguy.org/clicker} lets you mass send votes - it's quite fun to watch a confused professor.\\


\section*{Syllabus}  

\textit{A rigorous treatment of the concept of a limit, as applied to sequences, series and functions.}

\subsubsection*{Sequences} Real and complex sequences. Convergence, divergence and divergence to infinity. Sums and products of convergent sequences. The Sandwich Test. Sub-sequences, monotonic sequences, Bolzano-Weierstrass Theorem. Cauchy sequences and the general principle of convergence.

\subsubsection*{Series} Real and complex series. Convergent and absolutely convergent series. The Comparison Test for non-negative series and for absolutely convergent series. The Alternating Series Test.  Rearranging absolutely convergent series. Radius of convergence of power series. The exponential series.

\subsubsection*{Continuity} Limits and continuity of real and complex functions. Left and right limits and continuity. Maxima
and minima of real valued continuous functions on a closed interval. Inverse Function Theorem for
strictly monotonic real functions on an interval. 

\subsubsection*{Differentiability} An introduction to differentiability: definitions, examples, left and right derivative.

\subsection*{Appropriate books}

{\shortskip

K.~G.~ Binmore, \textit{Mathematical Analysis, A Straightforward Approach} (Cambridge University Press).\\

W. Rudin, \textit{Principles of Mathematical Analysis} (McGraw-Hill) (not recommended as an introduction)  
}

%\subsection*{Course Remarks}
%I typed-up the main bits during Easter as revision, and finished the rest over summer; it became quite apparent what he said was true - this stuff is  \emph{really} trivial in the scheme of things - so don't worry if Analysis first seems difficult; it's literally just messing with: 
%\begin{enumerate}
%\item $(a_n) \to a \iff \forall \epsilon > 0,~\exists N \in \mathbb{N} \text{ s.t. } |a_n - a| < \epsilon,~ \forall n \geq N$
%\item $\sum_{n=1}^{\infty} a_n = A \iff (s_n) = \sum_{k=1}^n a_k \to A$ 
%\item $f$ is continuous at $a \iff \forall \epsilon > 0,~\exists \delta > 0 \text{ s.t. } |x-a| < \delta \implies |f(x) - f(a)| < \epsilon$

%$\iff \exists \lim_{x \to a}f \iff \exists b: \forall \epsilon > 0,~\exists \delta > 0 \text{ s.t. } 0<|x-a| < \delta \implies |f(x) - b| < \epsilon $
%\item $f$ is differentiable at $a \iff \exists \lim_{x\to a} \frac{f(x) - f(a)}{x-a}$
%\end{enumerate}

}





\TableofContents
\input{Chap1.tex}
\input{Chap2.tex}		% Vector spaces
\input{Chap3.tex}		% Endomorhisms
\input{Chap4.tex}		% Endomorhisms


\end{document}