\documentclass[twoside]{scrartcl}

\usepackage[blacklinks]{drangreport}
% \usepackage[caecilia]{drangreport}

\usepackage{stmaryrd}

\usepackage{awlc-algebra}
\setcounter{section}{-1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%Messy stuff for plotting 
\newcommand{\vasymptote}[2][]{
    \draw [densely dashed,#1] ({rel axis cs:0,0} -| {axis cs:#2,0}) -- ({rel axis cs:0,1} -| {axis cs:#2,0});
}

\newcommand{\hasymptote}[2][]{
    \draw [densely dashed,#1] ({rel axis cs:0,0} -| {axis cs:0,#2}) -- ({rel axis cs:1,0} -| {axis cs:0,#2});
}


\pgfplotsset{
    double y domain/.code 2 args={
        \pgfmathsetmacro\doubleymin{#1*2}
        \pgfmathsetmacro\doubleymax{#2*2}
    },
    ejes/.style args={#1:#2 #3:#4}{
        double y domain={#3}{#4},
        domain=#1:#2,
        ymin=#3,ymax=#4, restrict y to domain=\doubleymin:\doubleymax,
        samples=60,
        enlargelimits=false,
        axis lines=middle,
        xticklabels=\empty, yticklabels=\empty,
        every axis plot post/.style={
            black,
            mark=none,
            smooth
        },
        scale only axis,
        width=6cm,
        height=6cm
    }
}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\title{Methods I}

\let\Im\relax
\DeclareMathOperator*{\Im}{Im}

\newcommand{\herm}[1]{\overline{#1}^\Trans}

\DeclareMathOperator*{\ch}{ch}
\renewcommand{\chi}{\textstyle\ch}

\begin{document}

\NotesTitle{}{Autumn 2014}{Mathematical Methods I}{Prof.~J.}{Mestel}
{
{\Large\bfseries\sffamily{Syllabus}}

\textit{The course supplies a firm grounding to A-level topics such as differentiation, integration, complex numbers and series expansions.
}



\begin{itemize}
\item Functions: Polynomial, rational, exponential, logarithmic, trigonometric and hyperbolic functions. Odd, even and inverse functions.

\item Limits: basic properties and evaluation. Continuity \& discontinuous functions.

\item Differentiation: First principles, differentiability; logarithmic and implicit differentiation; higher derivatives; Leibniz's formula; stationary points and points of inflexion; curve sketching; parametric representation, polar co-ordinates.

\item Power Series Expansions: The Mean Value Theorem. Taylor's Theorem with remainder. Infinite power series, radius of convergence. Ratio test; Taylor and Maclaurin Series. De l'Hopital's rule.
\itemIntegration: definition as Riemann limit; indefinite & definite integrals; the fundamental theorem of calculus; integration by substitution and by parts; partial fractions; Existence of improper and infinite integrals. Integrals over areas and volumes.
\item Complex Numbers: definition; the complex plane; standard and polar representation; de Moivre's Theorem; exp(z) and log(z)

\item First order differential equations. Separable, homogeneous and linear equations. Special cases. Linear higher order equations with constant coefficients,
\end{itemize}

Course content at \url{http://www.ma.ic.ac.uk/~ajm8/M1M1}\\

\subsection*{Appropriate books}

{\shortskip
M.~Liebeck \emph{A Concise Introduction to Pure Mathematics}.

K.~Houston \emph{How to Think Like a Mathematician}. 

E.~Hurst and M.~Gould \emph{Bridging the Gap to University Mathematics}. 

None are particularly recommended. The course is completely self-contained. But Stephenson's ``Maths for Science Students'', and Kreyszig's of a similar name are useful. 

}}

\TableofContents

\input{linalg-chap-01}		% Vector spaces
\input{linalg-chap-02}		% Endomorhisms


\end{document}